---
title: "Exploratory Data Analysis"
format: pdf
editor: visual
---

```{r include=F}
knitr::opts_chunk$set(warning=F, message=F, cache=T, echo=F, fig.height=3)
library(tidyverse)
library(ggplot2)
```

## Data Structure

We have data from 5 subjects. In the study each subject performs 8 different exercises while wearing 5 motion sensors on different parts of their body. Each sensor takes 9 measurements while the subject is performing the exercise. In all directions (x, y, and z) the sensor calculates the acceleration ($\frac{\text{meters}}{\text{second}^2}$), angular rate ($\frac{\text{radians}}{\text{second}}$), and magnetic field (relative).

$$
5 \text{ subjects } \times 8 \text{ exercises } \times 5 \text{ motion sensors } \times 9 \text{ measurements } = 1800 \text{ total data points}
$$

$$
8 \text{ exercises } \times 5 \text{ motion sensors } \times 9 \text{ measurements } = 360 \text{ data points per subject}
$$

$$
5 \text{ motion sensors } \times \text{ 9 measurements } = 45 \text{ data points per exercise}
$$

```{r}
data <- data.frame(
  Subject = rep(rep(rep(paste('Subject', seq(1:5), sep=' '),each=8),each=5),each=9),
  Exercise = rep(rep(paste('Exercise', seq(1:8), sep=' '),each=5),each=9),
  Sensor = rep(paste('Sensor', seq(1:5), sep=' '),each=9),
  Measurements = paste('Measurement', seq(1:9), sep=' ')
)
data %>% ggplot(aes(fill=Exercise, x=Subject)) +
  geom_bar() +
  labs(title='Data Decomposition Per Subject', y='Number of Data Points')
```

In our case, each data point is a time series of observations. The following is a single data point that represents acceleration in the $x$ direction ($\frac{\text{meters}}{\text{second}^2}$).

```{r}
tSesh <- read.table(file=paste('s1/e1/u2/template_session.txt', sep=''), header=TRUE, sep=';')
tSesh %>% ggplot() +
  geom_line(aes(x=time.index, y=acc_x)) +
  labs(x='Time Index', y='Acceleration in the x Direction', title='Session 1, Exercise 1, Sensor 2, Accelerometer (x)')
```

We can visualize a all of the measurements a sensor gives us:

```{r}
plottingFormat <- function(template_session) {
  acc <- template_session %>%
    select(time.index, acc_x, acc_y, acc_z) %>%
    pivot_longer(cols = starts_with('acc_'), 
                 names_to = 'coord', 
                 names_prefix = 'acc_', 
                 values_to = 'acc') 
  gyr <- template_session %>%
    select(time.index, gyr_x, gyr_y, gyr_z) %>%
    pivot_longer(cols = starts_with('gyr_'), 
                 names_to = 'coord', 
                 names_prefix = 'gyr_', 
                 values_to = 'gyr') 
  mag <- template_session %>%
    select(time.index, mag_x, mag_y, mag_z) %>%
    pivot_longer(cols = starts_with('mag_'), 
                 names_to = 'coord', 
                 names_prefix = 'mag_', 
                 values_to = 'mag') 
  list(acc, gyr, mag) %>% reduce(full_join, by=c('time.index', 'coord'))
}
plottingFormat(tSesh) -> p
p %>%
  ggplot() +
  geom_point(aes(x=time.index, y=acc, col=coord)) +
  facet_grid(coord ~., scales = "free") +
  labs(title='Session 1, Exercise 1, Sensor 2, Accelerometer',
       y='Acceleration',
       x='Time Index') +
  theme(legend.position = 'none')
p %>%
  ggplot() +
  geom_point(aes(x=time.index, y=gyr, col=coord)) +
  facet_grid(coord ~., scales = "free") +
  labs(title='Session 1, Exercise 1, Sensor 2, Gyroscope',
       y='Gyroscope',
       x='Time Index') +
  theme(legend.position = 'none')
p %>%
  ggplot() +
  geom_point(aes(x=time.index, y=mag, col=coord)) +
  facet_grid(coord ~., scales = "free") +
  labs(title='Session 1, Exercise 1, Sensor 2, Magnemometer',
       y='Magnemometer',
       x='Time Index') +
  theme(legend.position = 'none')
```

We can also visualize all of the data from a specific session and exercise:

```{r}
s <- 1 # Session number, 1:5
e <- 1 # Exercise, 1:8

tSesh1 <- read.table(file=paste('s',s,'/e',e,'/u',1,'/template_session.txt', sep=''), header=TRUE, sep=';') %>%
  plottingFormat() %>%
  rename(acc1=acc, gyr1=gyr, mag1=mag)
tSesh2 <- read.table(file=paste('s',s,'/e',e,'/u',2,'/template_session.txt', sep=''), header=TRUE, sep=';') %>%
  plottingFormat() %>%
  rename(acc2=acc, gyr2=gyr, mag2=mag)
tSesh3 <- read.table(file=paste('s',s,'/e',e,'/u',3,'/template_session.txt', sep=''), header=TRUE, sep=';') %>%
  plottingFormat() %>%
  rename(acc3=acc, gyr3=gyr, mag3=mag)
tSesh4 <- read.table(file=paste('s',s,'/e',e,'/u',4,'/template_session.txt', sep=''), header=TRUE, sep=';') %>%
  plottingFormat() %>%
  rename(acc4=acc, gyr4=gyr, mag4=mag)
tSesh5 <- read.table(file=paste('s',s,'/e',e,'/u',5,'/template_session.txt', sep=''), header=TRUE, sep=';') %>%
  plottingFormat() %>%
  rename(acc5=acc, gyr5=gyr, mag5=mag)
tSesh_all_sensors <- list(tSesh1, tSesh2, tSesh3, tSesh4, tSesh5) %>%
  reduce(full_join, by=c('time.index', 'coord'))

tTime <- read.table(file=paste('s',s,'/e',e,'/template_times.txt', sep=''), header=TRUE, sep=';')

colors <- c("Sensor 1" = "orange", "Sensor 2" = 'dodgerblue',
            "Sensor 3" = 'red2', "Sensor 4" = "green2",
            'Sensor 5' = 'black')
tSesh_all_sensors %>%
  ggplot() +
  geom_line(aes(x=time.index, y=acc1, color='Sensor 1')) +
  geom_line(aes(x=time.index, y=acc2, color='Sensor 2')) +
  geom_line(aes(x=time.index, y=acc3, color='Sensor 3')) +
  geom_line(aes(x=time.index, y=acc4, color='Sensor 4')) +
  geom_line(aes(x=time.index, y=acc5, color='Sensor 5')) +
  facet_grid(coord~.) +
  labs(title=paste('Subject ',s,', Exercise ',e,', Accelerometer', sep=''),
       x='Time Index',
       y='Acceleration',
       color='Sensor')

tSesh_all_sensors %>%
  ggplot() +
  geom_line(aes(x=time.index, y=gyr1), col='orange') +
  geom_line(aes(x=time.index, y=gyr2), col='dodgerblue') +
  geom_line(aes(x=time.index, y=gyr3), col='red2') +
  geom_line(aes(x=time.index, y=gyr4), col='green2') +
  geom_line(aes(x=time.index, y=gyr5)) +
  facet_grid(coord~., scales='free') +
  labs(title=paste('Subject ',s,', Exercise ',e,', Gyroscope', sep=''),
       x='Time Index',
       y='Gyroscope')
  
tSesh_all_sensors %>%
  ggplot() +
  geom_line(aes(x=time.index, y=mag1), col='orange') +
  geom_line(aes(x=time.index, y=mag2), col='dodgerblue') +
  geom_line(aes(x=time.index, y=mag3), col='red2') +
  geom_line(aes(x=time.index, y=mag4), col='green2') +
  geom_line(aes(x=time.index, y=mag5)) +
  facet_grid(coord~., scales='free') +
  labs(title=paste('Subject ',s,', Exercise ',e,', Magnetometer', sep=''),
       x='Time Index',
       y='Magnemometer')
```
